---
title: "Untitled"
date: "2026-09-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


You are consultant statisticians. A researcher has come to you to request help analyzing data generated by a designed experiment. They described the experiment in the following email: 

>I hope you are doing well. I am writing to request assistance with my statistical analysis for an swine nutrition trial. Please find a description of the study below.  
>
>The data were generated from an experiment that was studying the effect of different feed additives in swine production. A total of 300 pigs were used in this study, with dietary treatment applied to pen in a randomized complete block design.
>
>- Feed additives were applied at the pen level.
>- Observations: 3 pigs per pen were sampled on day 5, 10, and 17 and fecal sample was then placed in drying oven at 55 C for 48 hours to determine fecal dry matter.
>
>I need help with the following:  
>
>- Describing the experiment design for materials and methods. 
>- Analyzing the data and determining: 
  - If there are differences between feed additives on fecal DM, 
  - If there is a "better feed additive" (assume more DM is better). 


```{r}
url <- "https://raw.githubusercontent.com/stat870/fall2026/refs/heads/main/data/fecal_dm.csv"
dd_fecal <- read.csv(url)
dd_fecal$Pig <- as.factor(dd_fecal$Pig)
dd_fecal$Trt <- as.factor(dd_fecal$Trt)
dd_fecal$Room <- as.factor(dd_fecal$Room)
dd_fecal$Pen <- as.factor(dd_fecal$Pen)
dd_fecal$Day <- as.factor(dd_fecal$Day)
```

## EDA 


```{r}
dd_fecal |> 
  ggplot(aes(Day, Dry.matter...))+
  geom_boxplot()+
  facet_wrap(~Trt)+
  theme_classic()+
  theme(aspect.ratio = .5,
        panel.border = element_rect(color = "black", fill = NA))
```

## Statistical model


```{r}
library(lme4)
library(multcomp)

m1 <- lmer(Dry.matter... ~ Trt*Day + Room + (1|Room:Pen/Pig), data = dd_fecal)

(means <- emmeans(m1, ~Trt , at = list(Day = factor(17))))


cld(means)
```


