---
title: "Untitled"
author: "J Lacasa"
date: "2026-10-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(lme4)
library(emmeans)
```

```{r}
df <- read.csv("../data/cookies.csv")
df$Temperature_f <- as.factor(df$Temperature)
df$fat_perc <- as.factor(df$fat_perc)
df$Day <- as.factor(df$Day)
```

## Categorical predictor 


```{r}
df %>% 
  ggplot(aes(Temperature_f, force))+
  theme_classic()+
  scale_fill_manual(values = c("#DCD6F7", "#B4869F", "#985F6F", "#4E4C67"))+
  labs(x = "Temperature (°F)", fill = "Fat (%)", y = "Force (g)")+
  theme(aspect.ratio = .5)+
  geom_boxplot(aes(group = paste(Temperature, fat_perc),
                   fill = factor(fat_perc)))
```


The corresponding model is 
$$y_{ijk} = \mu + T_i + F_j + TF_{ij} + d_k + \varepsilon_{ijk} , \\
d_k \sim N(0, \sigma^2_d), \\
\varepsilon_{ijk} \sim N(0, \sigma^2_\varepsilon), $$

and can be fitted to the data with the following R code:

```{r}
m <- lmer(force ~ 1 + Temperature_f*fat_perc + (1|Day), data = df)
```

The degrees of freedom for the denominator are straightforward. 

```{r}
emmeans(m, ~Temperature_f:fat_perc, contr = list(c( 1, -1, rep(0, 10)), 
                                               c(0, 0, 0, 0, -1, 1, rep(0, 6))))
```


### t-test 

- **Objective**: compare treatment means between 375 degrees temperature and 4% fat and 400 degrees temperature and 4% fat. 
- Remember $t^\star = \frac{\hat\theta}{se(\hat\theta)}$ 
- $\hat\theta = \mu_{22} - \mu_{32}$ 
- $se(\hat\theta) = \sqrt{\frac{2 \sigma^2_\varepsilon}{r}}$

```{r}
d <- data.frame(Temperature = as.factor(c(375, 400)), 
                fat_perc = factor(4))
hat.mu <- predict(m, newdata = d, re.form = NA)
hat.theta <- hat.mu[2]-hat.mu[1]
```

```{r}
hat.sigma2 <- sigma(m)^2
hat.sigma2_b <- as.data.frame(VarCorr(m))[1,4]
reps <- n_distinct(df$Day)
se.hat.theta <- sqrt((2* hat.sigma2)/reps)
```

```{r}
emmeans(m, ~Temperature_f:fat_perc, contr = list(c(0, 0, 0, 
                                                 0, -1, 1, 
                                                 rep(0, 6))))$contrasts
```

```{r}
hat.theta
se.hat.theta

t.star <- hat.theta/se.hat.theta
t.star

dt(t.star, df = 33)
```

## Continuous predictor 

```{r}
df %>% 
  ggplot(aes(Temperature, force))+
  theme_classic()+
  scale_fill_manual(values = c("#DCD6F7", "#B4869F", "#985F6F", "#4E4C67"))+
  labs(x = "Temperature (°F)", fill = "Fat (%)", y = "Force (g)")+
  theme(aspect.ratio = .5)+
  geom_point(aes(group = paste(Temperature, fat_perc),
                 fill = factor(fat_perc)), shape =21, size =7 )
```



The corresponding model is 
$$y_{ijk} = \beta_{0j} + \beta_{1j} x_{ij} + d_k + \varepsilon_{ij} , \\
d_k \sim N(0, \sigma^2_d), \\
\varepsilon_{ij} \sim N(0, \sigma^2_\varepsilon), $$

and can be fitted to the data with the following R code:

```{r}
m <- lmer(force ~ Temperature*fat_perc + (1|Day), data = df)
```


```{r}
emmeans(m, ~Temperature_f:fat_perc, contr = list(c(0, 0, 0, 
                                                 0, -1, 1, 
                                                 rep(0, 6))))$contrasts
```

### Delta method 