---
title: "Untitled"
author: "J Lacasa"
date: "2026-10-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(lme4)
library(emmeans)
library(multcomp)
```

```{r}
df <- read.csv("../data/cookies.csv")
df$Temperature_f <- as.factor(df$Temperature)
df$fat_perc <- as.factor(df$fat_perc)
df$Day <- as.factor(df$Day)
```

## Categorical predictor 

A baker wanted to determine the effect that the amount of fat in a recipe of cookie dough would have on the texture of the cookie. 
She also wanted to determine if the temperature (°F) at which the cookies were baked would have an influence on the texture of the surface. 
The texture of the cookie is measured by determining the amount of force (g) required to penetrate the cookie surface. 

**Experimentation:** 

The process she used was to make a batch of cookie dough for each of the four recipes every day, and baked them one by one in the oven in different batches. 
She carried this process out each of four days when she baked cookies at three different temperatures. 

















```{r}
df %>% 
  ggplot(aes(Temperature_f, force))+
  theme_classic()+
  scale_fill_manual(values = c("#DCD6F7", "#B4869F", "#985F6F", "#4E4C67"))+
  labs(x = "Temperature (°F)", fill = "Fat (%)", y = "Force (g)")+
  theme(aspect.ratio = .5)+
  geom_boxplot(aes(group = paste(Temperature, fat_perc),
                   fill = factor(fat_perc)))
```


The corresponding model is 
$$y_{ijk} = \mu + T_i + F_j + TF_{ij} + d_k + \varepsilon_{ijk} , \\
d_k \sim N(0, \sigma^2_d), \\
\varepsilon_{ijk} \sim N(0, \sigma^2_\varepsilon), $$

and can be fitted to the data with the following R code:

```{r}
m <- lmer(force ~ 1 + Temperature_f*fat_perc + (1|Day), data = df)

DHARMa::simulateResiduals(m, plot = T)
```

The degrees of freedom for the denominator are straightforward. 

```{r}
means <- emmeans(m, ~Temperature_f:fat_perc)

cld(means, Letters = letters)
```

```{r}
emmeans(m, ~Temperature_f:fat_perc,
        contr = list(c(1, -1, rep(0, 10)), 
                     c(1, 0, 0, 1, 0, 0, -1, 0, 0, -1, 0, 0),
                     c(1/5,1/5,1/5,1/5,1/5, -1/7,-1/7,-1/7,-1/7,-1/7,-1/7,-1/7 )))
```


### t-test 

- **Objective**: compare treatment means between 375 degrees temperature and 4% fat and 400 degrees temperature and 4% fat. 
- Remember $t^\star = \frac{\hat\theta}{se(\hat\theta)}$ 
- $\hat\theta = \mu_{22} - \mu_{32}$ 
- $se(\hat\theta) = \sqrt{\frac{2 \sigma^2_\varepsilon}{r}}$

```{r}
d <- data.frame(Temperature = as.factor(c(375, 400)), 
                fat_perc = factor(4))
hat.mu <- predict(m, newdata = d, re.form = NA)
hat.theta <- hat.mu[2]-hat.mu[1]
```

```{r}
hat.sigma2 <- sigma(m)^2
hat.sigma2_b <- as.data.frame(VarCorr(m))[1,4]
reps <- n_distinct(df$Day)
se.hat.theta <- sqrt((2* hat.sigma2)/reps)
```

```{r}
emmeans(m, ~Temperature_f:fat_perc, contr = list(c(0, 0, 0, 
                                                 0, -1, 1, 
                                                 rep(0, 6))))$contrasts
```

```{r}
hat.theta
se.hat.theta

t.star <- hat.theta/se.hat.theta
t.star

dt(t.star, df = 33)
```

## Continuous predictor 

```{r}
df %>% 
  ggplot(aes(Temperature, force))+
  theme_classic()+
  scale_fill_manual(values = c("#DCD6F7", "#B4869F", "#985F6F", "#4E4C67"))+
  labs(x = "Temperature (°F)", fill = "Fat (%)", y = "Force (g)")+
  theme(aspect.ratio = .5)+
  geom_point(aes(group = paste(Temperature, fat_perc),
                 fill = factor(fat_perc)), shape =21, size =7 )
```



The corresponding model is 
$$y_{ijk} = \beta_{0j} + \beta_{1j} x_{ij} + d_k + \varepsilon_{ij} , \\
d_k \sim N(0, \sigma^2_d), \\
\varepsilon_{ij} \sim N(0, \sigma^2_\varepsilon), $$

and can be fitted to the data with the following R code:

```{r}
m <- lmer(force ~ Temperature+fat_perc+Temperature:fat_perc + (1|Day), data = df)

DHARMa::simulateResiduals(m, plot = T)
```


```{r}
emmeans(m, ~Temperature_f:fat_perc, contr = list(c(0, 0, 0, 
                                                 0, -1, 1, 
                                                 rep(0, 6))))$contrasts
```

### Delta method 


- The **delta method** sets the asymptotic variance of a function $g(\cdot)$ of the parameters ($g(\hat\beta)$): 
  $Var(g(\hat\beta)) \approx \nabla g(\hat\beta)^T Cov(\hat\beta) \nabla g(\hat\beta),$
  where $\nabla g(\hat\beta)$ is the gradient (vector of first derivatives) of function $g(\cdot)$ with respect to the parameters. 
- In this case $se(\hat{\beta}_1 + 2 \hat{\beta}_2 x_{ij}) = \sqrt(\widehat{Var}(\hat\beta_1)+ 4 x_{ij}^2\widehat{Var}(\hat\beta_2) + 4x_{ij}\widehat{Cov}(\hat\beta_2))$

```{r}
(covariance <- vcov(m1))
msm::deltamethod(g = ~ x2 + (2*x3*0), mean = coef(m1), cov = covariance)
emtrends(m1, ~ Total_N, "Total_N", at = list(Total_N = c(0, 100, 200)))
```

