# Linear mixed models  

September 2nd

## Announcements 

- Applied Linear Mixed Models Workshop 09/19-09/20. [Registration](https://www.k-state.edu/digital-ag-advanced-analytics/events/linear-mixed-models-2026.html) 
- [How to learn hard things](https://www.youtube.com/watch?v=RrLKXRXVrKM) 
- Read Chapter 2 of the Stroup et al. GLMM book (for this chapter only, 1ed is better than 2ed).  

**Semester project**: 

- Proposal due September 17th -- [see example 1](https://stat870.github.io/fall2026/assignments/proposal_example.pdf) | [see example 2](https://stat870.github.io/fall2026/assignments/proposal_example2.pdf)
- Must contain: Background, Objectives.  
- Data for the project must be ready to go. 
- Final product: Manuscript + Tutorial. 
- Working in teams vs. solo. 
- Start as soon as possible! 

## Review 

In what follows, we have a brief elaboration of what it means to model that "structure in the data" (i.e., the groups of similarly generated points). 

### Going from fixed effects to fixed+random effects  

The data below were generated by an experiment that tested 18 sorghum genotypes in a randomized complete block design. 

- Discuss the treatment structure 
- Discuss the design structure 

```{r}
dat_blocked <- agridat::omer.sorghum |> filter(env == "E3")

str(dat_blocked)
```

Now, remember that the blocks are supposed to indicate "groups of similar experimental units". 
Said "groups of similar experimental units" means that the assumption of independence between all observations (i.e., $y_{ij} = \mu_{0} + G_i + \varepsilon_{ij}$ and $\boldsymbol\varepsilon \sim N(\boldsymbol{0}, \sigma^2 \mathbf{I})$) would be kind of a stretch. 
In reality, all observations from the same block (are supposed to) have *something* in common. 
It's not reasonable to assume that the observations are independent, because observations from the same field have more in common than observations from different fields. 
They share more similar soil and, with that, a baseline fertility and yield. 

Basically, we expect the genotypes' relative performance to be similar across fields, but the baseline (a.k.a., the intercept) to be field-specific. Then, we could say    
$$y_{ij} = \beta_{0j} + G_i + \varepsilon_{ij}, \\ \varepsilon_{ij} \sim N(0, \sigma^2),$$  
where $\beta_{0j}$ is a block-specific intercept (i.e., $\beta_0 + b_j$, where $\beta_0$ is the overall intercept and $b_j$ is the block effect), and $G_i$ is the genotype effect.  
Now, there are different ways to model that block-specific intercept. 

This is a big forking path in statistical modeling. 
All-fixed models estimate the effects of **everything**. 
Mixed-effects models indicate *what is similar to what* via random effects. 

**Fixed effects** 

We could define an all-fixed model,  

$$y_{ij} = \beta_{0j} + G_i + \varepsilon_{ij}, \\ \beta_{0j} = \beta_0 + u_j \\ \varepsilon_{ij} \sim N(0, \sigma^2),$$  

where $u_j$ is the effect of the $j$th block on the intercept (i.e., on the baseline). 
In this case, $u_j$ is a fixed effect, which means it may be estimated via least squares estimation or maximum likelihood estimation. 
Under both least squares and maximum likelihood (assuming normal distribution), we may estimate the parameters by computing 

$$\hat{\boldsymbol{\beta}}_{ML} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y},$$

which yields the minimum variance unbiased estimate of $\boldsymbol{\beta}$. 

We still have $$\boldsymbol{\Sigma} = 
\begin{bmatrix} \sigma_{\varepsilon}^2 & 0 & 0 & 0 & \dots & 0 \\ 
0 & \sigma_{\varepsilon}^2 & 0 & 0 & \dots & 0 \\
0 & 0 & \sigma_{\varepsilon}^2 & 0 & \dots & 0 \\
0 & 0 & 0 & \sigma_{\varepsilon}^2 & \dots & 0 \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots\\  
0 & 0 & 0 & 0 & \dots & \sigma_{\varepsilon}^2 \end{bmatrix}$$. 


```{r}
m_fixed <- lm(yield ~ gen + rep, data = dat_blocked)

summary(m_fixed)
```

**Random effects** 

We could also assume that the effects of the $j$th block (i.e., $u_j$) arise from a random distribution. 
The most common assumption (and the default in most statistical software) is that 

$$u_j \sim N(0, \sigma^2_b).$$

Now, we don't estimate the effect, but the variance $\sigma^2_b$. 
Note that there are $J$ levels of the random effects, meaning that a random effect is always **categorical**.  
Also, now 

$$\hat{\boldsymbol{\beta}}_{REML} = (\mathbf{X}^T \mathbf{V}^{-1} \mathbf{X})^{-1}\mathbf{X}^T \mathbf{V}^{-1} \mathbf{y},$$

where $\mathbf{V} = Var(\mathbf{y})$ is the variance-covariance matrix of $\mathbf{y}$, 
including residual variance and random-effects variance. Note that this formula yields 
the same point estimate for $\boldsymbol{\beta}$, but with a different confidence interval.  

Now, the variance-covariance matrix $\boldsymbol{\Sigma}$ of the marginal distribution has changed. 

```{r}
library(lme4)
m_mixed <- lmer(yield ~ gen + (1|rep), data = dat_blocked)

summary(m_mixed)
```

## Generalities on mixed models 

Mixed models combine fixed effects and random effects. 
Generally speaking, we can write out a mixed-effects model using the model equation form, as   

$$\mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \mathbf{Z}\mathbf{u} + \boldsymbol{\varepsilon}, \\ 
\begin{bmatrix}\mathbf{u} \\ \boldsymbol{\varepsilon} \end{bmatrix} \sim \left(
\begin{bmatrix}\boldsymbol{0} \\ \boldsymbol{0} \end{bmatrix}, 
\begin{bmatrix}\mathbf{G} & \boldsymbol{0} \\
\boldsymbol{0} & \mathbf{R} \end{bmatrix} 
\right),$$

where $\mathbf{y}$ is the observed response, 
$\mathbf{X}$ is the matrix with the explanatory variables, 
$\mathbf{Z}$ is the design matrix,
$\boldsymbol{\beta}$ is the vector containing the fixed-effects, 
$\mathbf{u}$ is the vector containing the random effects, 
$\boldsymbol{\varepsilon}$ is the vector containing the residuals, 
$\mathbf{G}$ is the variance-covariance matrix of the random effects, 
and $\mathbf{R}$ is the variance-covariance matrix of the residuals. 
Note that $\mathbf{X} \boldsymbol{\beta}$ is the fixed effects part of the model, and 
$\mathbf{Z}\mathbf{u}$ is the random effects part of the model.


Using the probability distribution form, we can then say that $E(\mathbf{y}) = \mathbf{X}\boldsymbol{\beta}$ 
and $Var(\mathbf{y}) = \mathbf{V} = \mathbf{Z}\mathbf{G}\mathbf{Z}' + \mathbf{R}$. 
Usually, we assume $$\mathbf{G} = \sigma^2_u 
\begin{bmatrix} 1 & 0 & 0 & \dots 0 \\
0 & 1 & 0 & \dots & 0 \\
0 & 0 & 1 & \dots & 0 \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
0 & 0 & 0 & \dots & 1
\end{bmatrix}$$ and  $$\mathbf{R} = \sigma^2 
\begin{bmatrix} 1 & 0 & 0 & \dots & 0 \\
0 & 1 & 0 & \dots & 0 \\
0 & 0 & 1 & \dots & 0 \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
0 & 0 & 0 & \dots & 1
\end{bmatrix} $$. 

Then,   

$$\mathbf{y} \sim N(\boldsymbol{\mu}, \Sigma), \\ 
\Sigma = \begin{bmatrix} \sigma^2 + \sigma^2_u & \sigma^2_u & 0 & 0 & 0 & 0 &\dots & 0\\
\sigma^2_u & \sigma^2 + \sigma^2_u & 0 & 0 & 0 & 0 & \dots & 0 \\
0 & 0 & \sigma^2 + \sigma^2_u & \sigma^2_u  & 0 & 0 & \dots & 0 \\
0 & 0 & \sigma^2_u & \sigma^2 + \sigma^2_u  & 0 & 0 & \dots & 0 \\
0 & 0 & 0 & 0 & \sigma^2 + \sigma^2_u & \sigma^2_u  & \dots & 0 \\
0 & 0 & 0 & 0 & \sigma^2_u & \sigma^2 + \sigma^2_u  & \dots & \vdots \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \ddots & \vdots \\ 
0 & 0 & 0 & 0 & 0 & 0 & \dots & \sigma^2 + \sigma^2_u
\end{bmatrix}.$$

Take your time to digest the variance-covariance matrix above. What type of data do you think generated it? 

```{r echo=FALSE, message=FALSE, warning = FALSE}
Z <- getME(m_mixed, "Z")
sigma2_b <- 70
sigma <- 160

G <- diag(rep(sigma2_b^2, 4))

Z %*% G %*% t(Z) %>%
  as.matrix() %>% 
  as.data.frame() %>% 
  rownames_to_column("x") %>% 
  pivot_longer(cols = -c(x), names_to = "y") %>%
  mutate(x = as.numeric(x), y= as.numeric(y)) %>% 
  
  left_join(
  diag(sigma^2, nrow = nrow(dat_blocked)) %>% 
  as.data.frame() %>% 
  rownames_to_column("x") %>% 
  pivot_longer(cols = -c(x), names_to = "y") %>%
  mutate(y = str_replace(y, "V", "") %>% as.numeric(),
         x = as.numeric(x)) %>% 
  rename(s2 = value)) %>% 

  ggplot(aes(-x,y))+
  geom_tile(aes(fill = value+s2))+
  theme_minimal()+
  scale_fill_viridis_c()+
  labs(fill = "variance/covariance", 
       title = "Variance-covariance matrix\nof the marginal distribution \nfor the data in the\nsorghum block example.")+
  theme(aspect.ratio = 1, 
        axis.text = element_blank(), 
        legend.position = "bottom",
        panel.grid = element_blank(), 
        axis.title = element_blank(), 
        axis.ticks = element_blank())
```

### Random effects    

- By definition, random effects are regression coefficients that arise from a random distribution. 
- Typically, a random effect $u \sim N(0, \sigma^2_u)$.   
- Note that this model for the parameter may result in shrinkage. 
- We estimate the variance $\sigma^2_u$.  
- Calculating degrees of freedom can get much more complex than in all-fixed effects models (e.g., with unbalanced data, spatio-temporally correlated data, or non-normal data).  
- In the context of designed experiments, random effects are assumed to be independent to each other and independent to the residual.  


### Estimation of parameters   

"Estimation" is a term held mostly exclusive to fixed effects and variance components. 
Restricted maximum likelihood estimation (REML) is the default in most mixed effects models because, for small data (aka most experimental data), maximum likelihood (ML) provides variance estimates that are downward biased.
- In REML, the likelihood is maximized after accounting for the model’s fixed effects.  

- In ML, $\ell_{ML}(\boldsymbol{\sigma; \boldsymbol{\beta}, \mathbf{y}}) = - (\frac{n}{2}) \log(2\pi)-(\frac{1}{2}) \log ( \vert \mathbf{V}(\boldsymbol\sigma) \vert ) - (\frac{1}{2}) (\mathbf{y}-\mathbf{X}\boldsymbol{\beta})^T[\mathbf{V}(\boldsymbol\sigma)]^{-1}(\mathbf{y}-\mathbf{X}\boldsymbol{\beta})$  
- In REML, $\ell_{REML}(\boldsymbol{\sigma};\mathbf{y}) = - (\frac{n-p}{2}) \log (2\pi) - (\frac{1}{2}) \log ( \vert \mathbf{V}(\boldsymbol\sigma) \vert ) - (\frac{1}{2})log \left(  \vert \mathbf{X}^T[\mathbf{V}(\boldsymbol\sigma)]^{-1}\mathbf{X} \vert \right) - (\frac{1}{2})\mathbf{r}[\mathbf{V}(\boldsymbol\sigma)]^{-1}\mathbf{r}$, 
where $p = rank(\mathbf{X})$, $\mathbf{r} = \mathbf{y}-\mathbf{X}\hat{\boldsymbol{\beta}}_{ML}$.  
  - Start with initial values for $\boldsymbol{\sigma}$, $\tilde{\boldsymbol{\sigma}}$.  
  - Compute $\mathbf{G}(\tilde{\boldsymbol{\sigma}})$ and $\mathbf{R}(\tilde{\boldsymbol{\sigma}})$.  
  - Obtain $\boldsymbol{\beta}$ and $\mathbf{b}$.   
  - Update $\tilde{\boldsymbol{\sigma}}$.  
  - Repeat until convergence.  


### Fixed effects versus random effects  

What is behind a random effect:  

- $\hat{\boldsymbol{\beta}} \sim N \left( \boldsymbol{\beta}, (\mathbf{X}^T \mathbf{V}^{-1} \mathbf{X})^{-1} \right)$ 
- $u_j \sim N(0, \sigma^2_u)$ 
- What process is being studied?  
- How were the levels selected? (randomly, carefully selected)  
- How many levels does the factor have, vs. how many did we observe?   
- BLUEs versus BLUPs. 

Read more in in Gelman (2005, page 20), "Analysis of variance—why it is more important than ever" [[link](https://projecteuclid.org/journals/annals-of-statistics/volume-33/issue-1/Analysis-of-variancewhy-it-is-more-important-than-ever/10.1214/009053604000001048.full)], and [Gelman and Hill (2006), page 245](https://sites.stat.columbia.edu/gelman/arm/). 


### Standard errors 

- Blocks as fixed: 
  - $se(\hat{\mu_{i}}) = \sqrt{\frac{\sigma^2_\varepsilon}{\# blocks}}$
  - $se(\widehat{\mu_{i}-\mu_{i'}}) = \sqrt{\frac{2\sigma^2_\varepsilon}{\# blocks}}$

- Blocks as random: 
  - $se(\hat{\mu_{i}}) = \sqrt{\frac{\sigma^2_\varepsilon+\sigma^2_b}{\# blocks}}$
  - $se(\widehat{\mu_{i}-\mu_{i'}}) = \sqrt{\frac{2(\sigma^2_\varepsilon)}{\# blocks}}$


## Applied example: fungicide effects on barley genotypes   

Let's consider an experiment that aimed to study the effect of fungicide and barley genotypes on grain yield. The study includes two levels of fungicide, and 70 different genotypes, considering all their possible combinations. 

- What is the treatment structure of the experiment?
- Do we know the design structure of the experiment?


The statistical model for the split-plot can be described as 

$$y_{ijk} = \mu + F_i + G_j + F\times G_{ij} + b_k + w_{i(k)} + \varepsilon_{ijk}$$


**Now, the actual experiment was run like this:** 

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(agridat)
library(lme4)
library(emmeans)
```

```{r echo=TRUE}
df_fungicide <- agridat::durban.splitplot

df_fungicide %>% 
  ggplot(aes(bed, row))+
  coord_fixed()+
  geom_tile(aes(fill = fung))+
  geom_tile(aes(), color = "black", fill = NA)
```

- What is the treatment structure? 
- What are the experimental units? 
- How many independent observations per treatment factor? 
- What is the design structure? 


[Download R code](../scripts/09022026_splitplot.Rmd)

## For next week 

- Read Ch 4 in Johnson and Milliken 
- Read CH 2 in Stroup et al (2024)
